---
title: tergm tests
output:
  html_document:
    fig_width: 8
    highlight: kate
    theme: lumen
    toc: yes
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(cache=F, comment=NA, fig.align='center')
```

# Setup & data
```{r,  message=F}

# loadedNamespaces()
# unloadNamespace("ergm")
# unloadNamespace("statnet.common")

master_lib <- "~/R/GHmaster-library"

# #Make lib 1st position in libPath
# defaultPaths <- .libPaths()
# .libPaths(c(master_lib, defaultPaths))
# .libPaths()

#devtools::dev_mode(on=T, path = .libPaths()[1])
#library(network)
#library(ergm)
library(tergm, lib.loc = master_lib)
library(tsna)

sessioninfo::package_info(pkgs = c("statnet.common", "network", "rle",
                                   "ergm", "tergm"),
                          dependencies = FALSE)
```

# Types of networks {.tabset}

## first network

Standard network summary

```{r,  message=F}
data(samplk)
samplk1
```

## list

Replicates the standard summary for each network

```{r,  message=F}
# summary shows the nodes & edges at each slice
samp.list <- list(samplk1,samplk2,samplk3)
samp.list
```

## network.list

Only shows us that there are 3 network objects in this list

```{r,  message=F}
# summary only shows there are 3 network objects as components
samp.netlist <- network.list(samp.list)
samp.netlist
```

## NetSeries

Summary shows what `tergm` will send to `ergm` for term construction and estimation: the sum of the second and third neworks.

```{r,  message=F}
# summary shows nodes(2) + nodes(3), and edges(2) + edges(3)
samp.series <- NetSeries(samp.list)
samp.series
```

## networkDynamic

Summary shows the unique total node and edgecount (verified with the collapsed version).

```{r,  message=F}
# summary shows the unique # nodes and edges in both cases
# we start at 1 to make the series index consistent with the other
# network storage mechanisms

samp.dyn <- networkDynamic(network.list = samp.list, start=1)
samp.dyn
samp.collapse <- network.collapse(samp.dyn, onset=1, terminus=4)
samp.collapse
```

## Term counts for 2 panels: using base R

These replicate the edge counts performed by the operators, which makes it easier to verify.  But it only works for 2 panels.  Show the code for more info.

```{r}
form = network.edgecount(samplk1 | samplk2)
persist = network.edgecount(samplk1 & samplk2)
diss = -persist
dissolved = network.edgecount(samplk1) - persist
cross = network.edgecount(samplk2)
change = network.edgecount(xor(samplk1, samplk2))

cbind(form=form, persist=persist, diss=diss, dissolved=dissolved,
      cross=cross, change=change)
```

# Descriptives

Cross sectional stats and form/diss changes from tsna

```{r,  message=F}
samp.stats <- cbind(edges = tErgmStats(samp.dyn, "~edges"),
                    form = tEdgeFormation(samp.dyn),
                    diss = tEdgeDissolution(samp.dyn))

samp.stats
```

# Netstats and fits {.tabset}

These will be model-specific.  So we'll show results for different specifications.

## `Form + Diss` {.tabset .tabset-pills}

### summary_formula 

Works on NetSeries and nD (but not lists or netlists), and results are very different for these two, because they represent the network differently.

```{r,  message=F}
# summary.list <- summary(samp.list ~
#                             Form(~edges) +
#                             Diss(~edges))
# 
# summary.netlist <- summary(samp.netlist ~
#                             Form(~edges) +
#                             Diss(~edges))

summary.fd.series <- summary(samp.series ~
                        Form(~edges) +
                        Diss(~edges))
summary.fd.series

summary.fd.nD <- summary(samp.dyn ~
                        Form(~edges) +
                        Diss(~edges),
                      at=0:3)
summary.fd.nD
```

### fit 

Works for all data structures

```{r,  message=F}
fit.fd.list <- tergm(
  samp.list ~
    Form(~edges) +
    Diss(~edges),
  estimate = "CMLE",  times = c(1:3)
)

fit.fd.netlist <- tergm(
  samp.netlist ~
    Form(~edges) +
    Diss(~edges),
  estimate = "CMLE",  times = c(1:3)
)

fit.fd.series <- tergm(
  samp.series ~
    Form(~edges) +
    Diss(~edges),
  estimate = "CMLE" #,  times = c(1:3)
)

fit.fd.nD <- tergm(
  samp.dyn ~
    Form(~edges) +
    Diss(~edges),
  estimate = "CMLE",  times = c(1:3)
)
```

### Compare fit output

All give the same output.

```{r,  message=F}
cat("Coefs")
rbind(list=coef(fit.fd.list), 
      netlist = coef(fit.fd.netlist),
      netseries = coef(fit.fd.series),
      nD=coef(fit.fd.nD))

cat("nw.stats")
rbind(list=fit.fd.list$nw.stats, 
      netlist = fit.fd.netlist$nw.stats,
      netseries = fit.fd.series$nw.stats,
      nD= fit.fd.nD$nw.stats)

```

## `Form + Persist` {.tabset .tabset-pills}

### summary_formula 

Works on NetSeries and nD (but not lists or netlists), and results are very different for these two, because they represent the network differently.

```{r,  message=F}
# summary.list <- summary(samp.list ~
#                             Form(~edges) +
#                             Persist(~edges))
# 
# summary.netlist <- summary(samp.netlist ~
#                             Form(~edges) +
#                             Persist(~edges))

summary.fp.series <- summary(samp.series ~
                        Form(~edges) +
                        Persist(~edges))
summary.fp.series

summary.fp.nD <- summary(samp.dyn ~
                        Form(~edges) +
                        Persist(~edges),
                      at=0:3)
summary.fp.nD
```

### fit 

Works for all data structures

```{r,  message=F}
fit.fp.list <- tergm(
  samp.list ~
    Form(~edges) +
    Persist(~edges),
  estimate = "CMLE",  times = c(1:3)
)

fit.fp.netlist <- tergm(
  samp.netlist ~
    Form(~edges) +
    Persist(~edges),
  estimate = "CMLE",  times = c(1:3)
)

fit.fp.series <- tergm(
  samp.series ~
    Form(~edges) +
    Persist(~edges),
  estimate = "CMLE" #,  times = c(1:3)
)

fit.fp.nD <- tergm(
  samp.dyn ~
    Form(~edges) +
    Persist(~edges),
  estimate = "CMLE",  times = c(1:3)
)
```

### Compare fit output

All give the same output.

```{r,  message=F}
cat("Coefs")
rbind(list=coef(fit.fp.list), 
      netlist = coef(fit.fp.netlist),
      netseries = coef(fit.fp.series),
      nD=coef(fit.fp.nD))

cat("nw.stats")
rbind(list=fit.fp.list$nw.stats, 
      netlist = fit.fp.netlist$nw.stats,
      netseries = fit.fp.series$nw.stats,
      nD= fit.fp.nD$nw.stats)

```

## `Cross + Change` {.tabset .tabset-pills}

### summary_formula 

Works on NetSeries and nD (but not lists or netlists), and results are very different for these two, because they represent the network differently.

```{r,  message=F}
# summary.list <- summary(samp.list ~
#                             Form(~edges) +
#                             Persist(~edges))
# 
# summary.netlist <- summary(samp.netlist ~
#                             Form(~edges) +
#                             Persist(~edges))

summary.cc.series <- summary(samp.series ~
                        Cross(~edges) +
                        Change(~edges))
summary.cc.series

summary.cc.nD <- summary(samp.dyn ~
                        Cross(~edges) +
                        Change(~edges),
                      at=0:3)
summary.cc.nD
```

### fit 

Works for all data structures

```{r,  message=F}
fit.cc.list <- tergm(
  samp.list ~
    Cross(~edges) +
    Change(~edges),
  estimate = "CMLE",  times = c(1:3)
)

fit.cc.netlist <- tergm(
  samp.netlist ~
    Cross(~edges) +
    Change(~edges),
  estimate = "CMLE",  times = c(1:3)
)

fit.cc.series <- tergm(
  samp.series ~
    Cross(~edges) +
    Change(~edges),
  estimate = "CMLE" #,  times = c(1:3)
)

fit.cc.nD <- tergm(
  samp.dyn ~
    Cross(~edges) +
    Change(~edges),
  estimate = "CMLE",  times = c(1:3)
)
```

### Compare fit output

All give the same output.

```{r,  message=F}
cat("Coefs")
rbind(list=coef(fit.cc.list), 
      netlist = coef(fit.cc.netlist),
      netseries = coef(fit.cc.series),
      nD=coef(fit.cc.nD))

cat("nw.stats")
rbind(list=fit.cc.list$nw.stats, 
      netlist = fit.cc.netlist$nw.stats,
      netseries = fit.cc.series$nw.stats,
      nD= fit.cc.nD$nw.stats)

```

## Bare terms (no operators) {.tabset .tabset-pills}

### summary_formula 

Works on NetSeries and nD (but not lists or netlists), and results are very different for these two, because they represent the network differently.

```{r,  message=F}
# summary.list <- summary(samp.list ~
#                             Form(~edges) +
#                             Persist(~edges))
# 
# summary.netlist <- summary(samp.netlist ~
#                             Form(~edges) +
#                             Persist(~edges))

summary.bare.series <- summary(samp.series ~ edges)
summary.bare.series

summary.bare.nD <- summary(samp.dyn ~ edges,
                          at=0:3)
summary.bare.nD
```

### fit 

Works for all data structures

```{r,  message=F}
fit.bare.list <- tergm(
  samp.list ~edges,
  estimate = "CMLE",  times = c(1:3)
)

fit.bare.netlist <- tergm(
  samp.netlist ~edges,
  estimate = "CMLE",  times = c(1:3)
)

fit.bare.series <- tergm(
  samp.series ~ edges,
  estimate = "CMLE" #,  times = c(1:3)
)

fit.bare.nD <- tergm(
  samp.dyn ~ edges,
  estimate = "CMLE",  times = c(1:3)
)
```

### Compare fit output

All give the same output.

```{r,  message=F}
cat("Coefs")
rbind(list=coef(fit.bare.list), 
      netlist = coef(fit.bare.netlist),
      netseries = coef(fit.bare.series),
      nD=coef(fit.bare.nD))

cat("nw.stats")
rbind(list=fit.bare.list$nw.stats, 
      netlist = fit.bare.netlist$nw.stats,
      netseries = fit.bare.series$nw.stats,
      nD= fit.bare.nD$nw.stats)

```

# Summary

**The simple descriptives:**

```{r,  message=F}
samp.stats

test <- rbind(names(coef(fit.cc.nD)), round(coef(fit.cc.nD), 3))
```
____

**The model stats for each model**

model | names | nw.stats
:---- | :--- | :---
Form + Diss:  | `r names(fit.fd.nD$nw.stats)` | `r fit.fd.nD$nw.stats`  
Form + Persist: | `r names(fit.fp.nD$nw.stats)` | `r fit.fp.nD$nw.stats`  
Cross + Change: | `r names(fit.cc.nD$nw.stats)` | `r fit.cc.nD$nw.stats`  
Bare, no operators:  | `r names(fit.bare.nD$nw.stats)` | `r fit.bare.nD$nw.stats`

____

**The model coefs for each model**

model | names | coefs
:---- | :--- | :---
Form + Diss: | `r names(coef(fit.fd.nD))` | `r round(coef(fit.fd.nD), 3)`  
Form + Persist: | `r names(coef(fit.fp.nD))` | `r round(coef(fit.fp.nD), 3)`  
Cross + Change: | `r names(coef(fit.cc.nD))` | `r round(coef(fit.cc.nD), 3)`  
Bare, no operators: | `r names(coef(fit.bare.nD))` | `r round(coef(fit.bare.nD), 3)`


Note that:

The `Form + Persist` coefs are: `r round(coef(fit.fp.nD), 3)`

The `Cross + Change` coefs are:  `r round(coef(fit.cc.nD), 3)`

By inspection and implication:


* `Form = Cross + Change`:  -2.498 = -0.896 + (-1.602)

`Form` operates on the union of $Y_t$ and $Y_{t+1}$.  `Cross` operates on $Y_{t+1}$, `Change` operates on the ties formed, and the ties dissolved (xor($Y_t, Y_{t+1}$)).  So taking the union of `Cross` with `Change` is  $Y_{t+1}$ plus the dissolved ties, the union of $Y_t$ and $Y_{t+1}$

    * Cross = Form - Change

* `Persist = Cross - Change`:  0.707 = -0.896 - (-1.602)

Persist operates on the intersection
 

And from Pavel:

When there are 3 time points, both `Cross()` and bare terms will not "see" the first network. t1 network will only affect `Change(), Form(), and Diss()/Persist().`

# Simulation {tabset}

We'll just use the fit from the `NetSeries` object

## Separable model
```{r, message=F, warning=F}
sep.sim <- simulate(fit.fd.series, nsim = 1, 
                    time.slices = 1000, 
                    nw.start = "first",
                    monitor = ~edges,
                    stats = T,
                    output = "stats")
```

#### Monitor stats: edges
```{r, message=F, warning=F}

target = summary(samp.series ~ edges)

cbind(obs = target,
      sim = colMeans(sep.sim$stats))

plot(as.numeric(sep.sim$stats), type="l",
     main = "Cross-sectional edgecount in simulated network timeseries",
     sub = "Model: Form(~edges + mutual) + Diss(~edges + mutual)",
     ylab="edges") 
abline(h=target, col="red")

```

# Add mutuals

## Summary 

```{r, message=F, warning=F}
summary(samplk1 ~edges + mutual)
summary(samp.series ~
          Cross(~edges + mutual) +
          Change(~edges + mutual) +
          Form(~edges + mutual) +
          Diss(~edges + mutual))

```

## Fit {.tabset}

### Separable
```{r, message=F, warning=F}
sep.fit <- tergm(
  samp.series ~
    Form(~edges + mutual) +
    Diss(~edges + mutual),
  estimate = "CMLE"
)

summary(sep.fit)
```

### Joint
```{r, message=F, warning=F}
jnt.fit <- tergm(
  samp.series ~
    Cross(~edges + mutual) +
    Change(~edges + mutual),
  estimate = "CMLE"
)

summary(jnt.fit)
```

## Simulation {.tabset }

### Separable

```{r, message=F, warning=F}
sep.sim <- simulate(sep.fit, nsim = 1, 
                    time.slices = 1000, 
                    nw.start = "first", 
                    monitor = ~edges,
                    stats = T,
                    output = "stats")
```

#### Monitor stats: edges
```{r, message=F, warning=F}

target = summary(samp.series ~ edges)

cbind(obs = target,
      sim = colMeans(sep.sim$stats))

plot(as.numeric(sep.sim$stats), type="l",
     main = "Cross-sectional edgecount in simulated network timeseries",
     sub = "Model: Form(~edges + mutual) + Diss(~edges + mutual)",
     ylab="edges") 
abline(h=target, col="red")

```

#### Model stats
```{r, message=F, warning=F}
cbind(obs = sep.fit$nw.stats,
      sim = colMeans(sep.sim$stats.gen),
      ratio = with(data.frame(obs = sep.fit$nw.stats,
                              sim = colMeans(sep.sim$stats.gen)),
                     obs/sim))

plot(sep.sim$stats.gen)

```



### Joint

```{r, message=F, warning=F}
jnt.sim <- simulate(sep.fit, nsim = 1, 
                    time.slices = 1000, 
                    nw.start = "first", 
                    monitor = ~edges,
                    stats = T,
                    output = "stats")
```

#### Monitor stats: edges
```{r, message=F, warning=F}
cbind(obs = summary(samp.series ~ edges),
      sim = colMeans(jnt.sim$stats))

target = summary(samp.series ~ edges)
plot(as.numeric(jnt.sim$stats), type="l",
     main = "Cross-sectional edgecount in simulated network timeseries",
     sub = "Model: Cross(~edges + mutual) + Change(~edges + mutual)",
     ylab="edges") 
abline(h=target, col="red")

```

#### Model stats

```{r eval=F}
cbind(obs = jnt.fit$nw.stats,
      sim = colMeans(jnt.sim$stats.gen),
      ratio = with(data.frame(obs = jnt.fit$nw.stats,
                              sim = colMeans(jnt.sim$stats.gen)),
                     obs/sim))

plot(jnt.sim$stats.gen)
```


