---
title: "Extending ERGM Functionality within statnet: Building Custom User Terms"
author: statnet development team
date: Sunbelt 2019
output: beamer_presentation
---

## Outline

- Quick review of ERGMs
- Overview of change statistics
- Overview of steps for writing one’s own statistic
- Detailed look at the R code
- Detailed look at the C code
- Group exercise

```{r, LoadErgm, fig.cap="A network", echo=FALSE, message=FALSE}
library(ergm)
```

---

## ERGM basic expression

Probability of observing a network (set of relationships) y on a given set of actors: 
\[
P(Y=y) = \frac{ \exp \{ \theta^\top g(y)\} }{k(\theta)},
\]
where

- $g(y) =$ vector of network statistics (like the predictors in a standard
linear regression)
- $\theta =$ vector of model parameters (like the coefficients in a linear regression) 
- $k(\theta) =$ normalizer, equal to the sum for all possible $y^*$ of 
$\exp\{ \theta^\top g(y^*)\}$.

<!--Bahadur (1961), Besag (1974), Frank (1986); Wasserman and Pattison (1996) -->

---

## ERGM logit formulation

The statement 
\[
P(Y=y) = \frac{ \exp \{ \theta^\top g(y)\} }{k(\theta)}
\]

implies that 
\[
\log \frac{P(Y_{ij}=1 \, |\, y_{ij}^c)}{P(Y_{ij}=0 \, |\, y_{ij}^c)} =
\theta^\top \left[ g(y_{ij}^+) - g(y_{ij}^-) \right] = 
\theta^\top \delta(y)_{ij},
\]
where

- $Y_{ij}$ is the (random) value of the tie between $i$ and $j$
- $y_{ij}^c$ is the whole network $y$ except for the value of $y_{ij}$ itself
- $y_{ij}^+$ is the whole network $y$ with $y_{ij}$ set to 1
- $y_{ij}^-$ is the whole network $y$ with $y_{ij}$ set to 0
- $\delta(y)_{ij} = g(y_{ij}^+) - g(y_{ij}^-)$ is the vector of change statistics,
that is, the difference between the $g(y)$ vectors evaluated at two networks that
differ only in the $i,j$ entry.


---

```{r, net, fig.cap="A network", echo=FALSE}
data(faux.mesa.high)
plot(faux.mesa.high)
```

